Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
paper s5 conflics solved

In this section we describe our main contribution: Unbalanced Mallows model
(UMM), a novel algorithm for the optimization of an expensive black-box function over
permutation spaces. This algorithm is a probabilistic method based on the
Mallows model and uBorda learning.\MANUEL{Perhaps we should call it UMM Algorithm? otherwise people would complain. Or Unbalanced Mallows Algorithm (UMA)?}

\EKHINE{10. In Algorithm 1 'S' and 'F' are *sets* (thus, unordered). Therefore, when adding sigma and f(sigma) the correspondence between the two elements seems to be lost. This mapping would be better represented as a set of *pairs* < sigma, f(sigma) >. You always add these elements to both sets at the same time in step 8, anyway. MANUEL: FIXED
}

@@

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are generated uniformly at random  (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to learn a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). 
Together with the parameter $\theta$, the algorithm defines the Mallows model $MM(\hat\sigma_0, \theta)$. 
A decreasing $\theta$ parameter along iterations (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. 
Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.
\EKHINE{(3) In section 3, second half of the second paragraph (from "Then, we sample a new candidate [...]"): the algorithm 1 is not properly described. Please provide a description in line with algorithm 1 and, above all, describe the rationale behind the choice of line 7 of algorithm 1.
I like it, but it deserves to be better described. It seems to me a sort of "novelty-driven" optimization.(4) A small detail, but requires a clarification. In page 6, learning step paragraph: the weight formula uses fitness in the exponent, while the base rho is in (0,1). Clearly the authors are working with minimization problems and want to increase the weight when fitness improves/decreases. This is ok, but this only happens with their formula when fitnesses are greater than 1. I know this is quite usual in the problems analyzed by the authors, but it may not be so usual in general. Please add a clarification.}

@@

As a consequence of this important distinction, when dealing with \manuel{data in the form of}{} permutations, in particular in 
combinatorial problems, we must convert rankings to orderings by inverting the
permutation, as done in Eq.~\ref{eq:lop} and Eq.~\ref{eq:pfsp}, for the LOP and
PFSP, respectively. This might seem an implementation detail but a failure to use the correct interpretation has grave consequences.

@@

\myparagraph{Convergence analysis of UMM} We start our analysis with the three
configurations of UMM. We can see a common trend in all instances: After
the 10 random initial permutations, the fitness of new solutions decreases
with a linear trend along iterations. The (close to) linear decrease is given
by the linear decrease in the expected distance\MANUEL{distance to the sampling model or variance of the sampling model?} of the sampling model, as described in Section~\ref{sec:sampling}. Overall, the rate of improvement
does not seem to converge, not even for the longer runs of 400 evaluations. Therefore, we conjecture that longer runs are likely to produce further improved solutions. Moreover, a non-linear decrease in the expected distance should result in a faster convergence for larger evaluation budgets. \manuel{The
average runtime for all the instances is around 0.3 minutes, 1.15 minutes and
4.5 minutes respectively for the versions with 100, 200 and 400 fitness
evaluations.}{}\MANUEL{We should discuss the time when discussing the tables.}

@@

\myparagraph{Comparison of UMM and CEGO}

When comparing the UMM and CEGO configurations shown in Fig.~\ref{fig:results},
we can observe that all UMM configurations perform much better than the
original CEGO (\CEGOorig), while they converge slower than our modified
\CEGOinv. However, the rate of improvement of the UMM variants does not seem to
slow down near the maximum budget given and, in several instances such as
\texttt{rec31} and \texttt{N-sgb75-01}, matches and surpasses the solutions
found by \CEGOinv. Interestingly, this is the case, even for the shorter budget
configurations of UMM, i.e., UMM$_{100}$ and UMM$_{200}$. We also observe a
pattern that UMM performs better, or \CEGOinv performs worse, on larger
instances, as can be seen by comparing \texttt{rec19} with \texttt{rec31}.

@@

Moreover, UMM seems to scale good for larger instances in quality performance. Note that the instances are order from top to bottom in the size of the permutations. 
The computational complexity of UMM is \ekhine{}{ $O(n \log n)$ }at each iteration while both versions of CEGO need \manuel{tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p50-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p50-01}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

@@

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\begin{table*}[tb]
  \caption{Mean fitness (and standard deviation) of the best solution found. % at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of UMM minus the one of CEGO. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
  \resizebox{\textwidth}{!}{%
\begin{tabular}{r*{5}{r}r@{\hskip 0.5ex}l*{2}{r}}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-p50-02     & 19090.5                      & (285.4)                                                                         & 21746.6   & (645.4)   & {[}-3140.6,& -2171.6{]}    & 22.9 & 0.5 \\
         & N-p50-01     & 18197.7                      & (225.6)                                                                         & 21161.8   & (657.4)   & {[}-3447.4,& -2480.8{]}    & 23.2 & 0.5 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}}
\end{table*}

>>>>

In this section we describe our main contribution: Unbalanced Mallows model
(UMM), a novel algorithm for the optimization of an expensive black-box function over
permutation spaces. This algorithm is a probabilistic method based on the
Mallows model and uBorda learning.%\MANUEL{Perhaps we should call it UMM Algorithm? otherwise people would complain. Or Unbalanced Mallows Algorithm (UMA)?}

@@

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are generated uniformly at random  (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to learn a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). 
Together with the parameter $\theta$, the algorithm defines the Mallows model $MM(\hat\sigma_0, \theta)$. 
A decreasing $\theta$ parameter along iterations (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. 
Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}, also  in  Section~\ref{sec:sampling}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.
%\EKHINE{(3) In section 3, second half of the second paragraph (from "Then, we sample a new candidate [...]"): the algorithm 1 is not properly described. Please provide a description in line with algorithm 1 and, above all, describe the rationale behind the choice of line 7 of algorithm 1.
%I like it, but it deserves to be better described. It seems to me a sort of "novelty-driven" optimization.(4) A small detail, but requires a clarification. In page 6, learning step paragraph: the weight formula uses fitness in the exponent, while the base rho is in (0,1). Clearly the authors are working with minimization problems and want to increase the weight when fitness improves/decreases. This is ok, but this only happens with their formula when fitnesses are greater than 1 NOOO. I know this is quite usual in the problems analyzed by the authors, but it may not be so usual in general. Please add a clarification.}

@@

As a consequence of this important distinction, when dealing with  permutations, in particular in 
combinatorial problems, we must convert rankings to orderings by inverting the
permutation, as done in Eq.~\ref{eq:lop} and Eq.~\ref{eq:pfsp}, for the LOP and
PFSP, respectively. This might seem an implementation detail but a failure to use the correct interpretation has grave consequences.

@@

\myparagraph{Convergence analysis of UMM} We start our analysis with the three
configurations of UMM. We can see a common trend in all instances: After
the 10 random initial permutations, the fitness of new solutions decreases
with a linear trend along iterations. The (close to) linear decrease is given
by the linear decrease in the expected distance of the model, as described in Section~\ref{sec:sampling}. Overall, the rate of improvement
does not seem to converge, not even for the longer runs of 400 evaluations. Therefore, we conjecture that longer runs are likely to produce further improved solutions. Moreover, a non-linear decrease in the expected distance should result in a faster convergence for larger evaluation budgets. \manuel{The
average runtime for all the instances is around 0.3 minutes, 1.15 minutes and
4.5 minutes respectively for the versions with 100, 200 and 400 fitness
evaluations.}{}\MANUEL{We should discuss the time when discussing the tables.}
\EKHINE{luego el detalle, que la gente no llega al final }

@@

\myparagraph{Comparison of UMM and CEGO}
When comparing the UMM and CEGO configurations shown in Fig.~\ref{fig:results},
we can observe that all UMM configurations perform much better than the
original CEGO (\CEGOorig), while they converge slower than our modified
\CEGOinv. However, the rate of improvement of the UMM variants does not seem to
slow down near the maximum budget given and, in several instances such as
\texttt{rec31} and \texttt{N-sgb75-01}, matches and surpasses the solutions
found by \CEGOinv. Interestingly, this is the case, even for the shorter budget
configurations of UMM, i.e., UMM$_{100}$ and UMM$_{200}$. We also observe a
pattern that UMM performs better, or \CEGOinv performs worse, on larger
instances, as can be seen by comparing \texttt{rec19} with \texttt{rec31}.

@@

Moreover, UMM seems to scale good for larger instances in quality performance. Note that the instances are order from top to bottom in the size of the permutations. 
The computational complexity of UMM is $O(n \log n)$ at each iteration while both versions of CEGO need \manuel{tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p50-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

@@

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\begin{table*}[tb]
  \caption{Mean fitness (and standard deviation) of the best solution found. % at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of UMM minus the one of CEGO. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
  \resizebox{\textwidth}{!}{%
\begin{tabular}{r@{\hskip -1ex}r*{2}{r@{\hskip 1ex}l}r@{\hskip 0.5ex}lr@{\hskip 0.75ex}r}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-p50-02     & 19090.5                      & (285.4)                                                                         & 21746.6   & (645.4)   & {[}-3140.6,& -2171.6{]}    & 22.9 & 0.5 \\
         & N-p50-01     & 18197.7                      & (225.6)                                                                         & 21161.8   & (657.4)   & {[}-3447.4,& -2480.8{]}    & 23.2 & 0.5 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}}
\end{table*}
